Species motif participation provides unique information about species risk of extinction

### 2022-06-17 manuscript/response_PRSB_round2.tex

        In addition, we believe that the Reviewer has over-estimated our reliance on the BioRxiv pre-print currently under review. Our previous draft had only three references to this pre-print: when describing the generation of simulated networks, when noting that degree and motif participation are not independent (which we also demonstrate \emph{de novo} in the present manuscript), and when briefly comparing the dynamical and Bayesian results in the discussion.
        We have now removed two of these references.
        We present the full network-generation methods in \emph{Appendix 1} and, to avoid confusion by referring to results presented in the preprint, have softened our language to say that degree and motif participation \emph{may} be correlated. 
        The latter is not a major point in our manuscript; we mention it only to emphasize that controlling for the total number of motifs in which a species appears does not necessarily control for degree. 
        Thus, we see some overlap in the associations of persistence with degree and with normalised motifs.

        We have retained the reference to the pre-print in the discussion because this brief comparison of results using dynamical and Bayesian simulations appears to be exactly what the Reviewer has asked for.
        As the Reviewer points out, different approaches to simulating disturbance can have different results, and we explicitly mention the main differences in the Discussion. 
        As the pre-print is freely available to any who wish to critically examine it, we believe that this is a good compromise between maintaining conciseness of the present manuscript and comparing different simulation approaches.
        If, all going well, the preprint is accepted for publication by the time the present manuscript is accepted, we will of course update the reference to provide the peer-reviewed version of the text.

@@

        \them
        Line 247-254: This would all be greatly enhanced if there were some error bars on the figures.

        \us We have added error bars representing 95\% confidence intervals to this figure.

        \them
        Line 259-262: This result seems odd to me. Is this just because all of the potential apparent competition interactions are filled first, and really is more of an artefact of the constraints about the types of motifs that are possible within these networks? And perhaps the connectances being explored are unrealistic given these constraints?

        \us While we do not use a generative model that fills some motifs/interactions before others, we do not see any particular reason why apparent competition motifs should be filled first. More likely, the high frequencies of apparent competition reflects the tendency for resources to outnumber consumers, creating the classic ``pyramid shape'' food web. We can confirm, however, that the connectances we explore are realistic. We compare the structure of our simulated networks to the largest set of high-quality, similarly-sized empirical networks we could find. These networks had 50-77 species and connectances of 0.095-0.227, covering the upper half of connectances we simulate. We also, per the Associate Editor's request, compare the motif profiles and motif participation of these empirical webs to our simulated webs. We hope that this comparison will demonstrate the realism of our simulated networks.

### 2022-06-17 manuscript/reorder_main_text.tex

    Bottom-up effects of disturbances to basal resources can strongly affect the persistence of consumers, but it is difficult to predict which species will be most affected by increasing disturbance.
    Coarse measures of a species' structural role, like in-degree and trophic level, are known to affect a consumer's vulnerability to bottom-up disturbance.
    We are interested in whether a richer set of species roles based on participation in three-species motifs is also related to persistence, and how these motif-based roles relate to simpler measures.
    We show that a consumer species' participation in different motifs is related to its probability of persistence. 
    Similarly, the overall motif profile of a network was related to the average persistence of consumer species.
    Notably, we found that networks containing a higher proportion of the `omnivory' motif tend to have lower mean persistence following disturbance, while more frequent participation in the omnivory motif was associated with higher species-level probabilities of persistence.
    Additionally, we find that relationships between motif participation and persistence can synthesize the information provided by in-degree and trophic level.
    Motif participation therefore captures important trends in species persistence and provides a rich description of species' structural roles in their communities.

@@

    %\subsection*{1 Why are bottom-up effects important for understanding stability?}
     The plant community is the foundation upon which the myriad species in food webs depend for their survival.
     Disturbances in the plant community have been shown to affect important ecosystem properties such as primary \citep{Hector1999} and secondary production \citep{borer2012plant}, soil respiration and carbon cycling \citep{chen2019plant} and consumer diversity \citep{scherber2010bottom, Baiser2016}.
     As such, the structure of the plant community affects stability across all levels of ecosystem organization, which is crucial for sustaining a healthy energy flow in the ecosystem as a whole~\citep{proulx2010diversity, scherber2010bottom, Rosenblatt2016}. 
     For example, the loss of diversity at the basal level typically causes declines in both abundance and richness of all types of consumers: herbivores, predators, parasitoids, etc. \citep{scherber2010bottom}.

    %\subsection*{2 Global and local network properties are known to affect stability}
    How disturbances, such as population decline or extinction, lead to loss of other species via direct and indirect effects (i.e., secondary extinctions) has for long been vibrant area of research \citep{Santos2021,curtsdotter2011robustness, dunne2009cascading, Eklof2006}.
    Several global (whole-network) as well as local (single-species) properties are known to influence how species respond to disturbances. Two well-known, influential global properties are network size, the number of species present in the network, and network connectance, the number of realized feeding interactions.
    Based on simulations of random networks, large networks were initially thought to be less stable~\citep{May1972}.
    The indisputable presence of large, stable, empirical networks has spurred research into how other structural characteristics, including connectance, may allow large numbers of species to persist~\citep{Dunne2002}.
    Highly-connected networks may be more resistant to secondary extinctions because consumers, on average, have access to a larger number of prey species and can switch from a declining or extinct prey to another resource~\citep{Dunne2002, Eklof2006,Baumgartner2015}. However, a well-connected network also includes more pathways through which a disturbance can spread~\citep{Dunne2004, Vieira2015}. Depending on the strength and arrangements of interactions, this means that species may be impacted through several paths, amplifying the original disturbance and causing more secondary extinctions.
    The net effect of connectance on extinction therefore likely depends on finer-scale (species-level) network structure, as well as the type and severity of disturbance.

    Species-level properties known to affect species' response to disturbances include trophic level (a measure of the path length between a consumer and a basal resource) and in-degree (number of prey items of a predator). 
    These can also be referred to as local properties of the network.
    Species with high trophic levels (long paths to basal resources) and low in-degree (few prey) are generally at higher risks of secondary extinction \citep{binzer2011susceptibility, Eklof2006}, and disturbances to plant populations cause particularly many econdary extinctions \citep{Curtsdotter2011}. %including secondary extinction due to reduced plant populations.
    However, in-degree and trophic level are both relatively coarse descriptions of a species' role in its food web~\citep{Cirtwill2018FoodWebs}. 
    A more detailed perspective is desirable since, as well as the properties of the consumer itself, a consumer's risk of secondary extinction after a disturbance to plants depends on the extinction risk of its prey.
    For example, a generalist consuming many specialist (and therefore high-risk) prey may be at greater risk of extinction than a specialist which consumes a highly generalist resource that is more unlikely to go extinct (Fig.~\ref{fig:concept}).

@@

        A middle‐ground approach for simulating secondary extinctions is to use Bayesian networks \citep{Eklof2013,Haussler2020}. 
        In this framework, a consumer's probability of extinction depends on the fraction of resources lost ($f = k/n$ for a species with $n$ resources, of which $k$ have gone extinct) and on a baseline probability of extinction ($\pi$) which captures factors affecting the focal species' extinction probability that are not related to the network structure (e.g., disease, stochastic extinction of small populations).
        Modelling Bayesian networks is fast and computationally efficient, allowing analysis of networks with high species richness \citep{Haussler2020}. 
        This increased speed allows us to simulate a range of disturbance severity rather than single species removals as in many previous studies (e.g.,~\citealp[]{Memmott2004,Staniczenko2010,Dunne2004}).
        Moreover, Bayesian network simulations are less parameter-intensive than dynamical models, meaning that they are less sensitive to assumptions made in the modelling process~\citep{Eklof2013}.

@@

		% how we calculate persistences 
        Beginning with basal resources and working systematically up through the food web, we use Equation \ref{betafunc} to calculate each species' $i$ probability of extinction $P(\lnot i|f)$.
        Using this probability, we then perform Bernoulli trials to determine whether the species goes extinct or not. 
        Species are considered extinct if a random number drawn from a uniform distribution $[0-1]$ is less than $P(\lnot i|f)$.
        These simulated extinctions are used when calculating the fraction of resources lost $f$ for each consumer.
        This allows us to update Equation~\ref{betafunc} with information on extinctions of a focal consumer's resources and simulate bottom-up effects. As this method is probabilistic, we repeat these calculations 100 000 times per network, with unique random draws.
	    For each species, we then define probability of persistence as the fraction of simulations in which the species persists. There are methods for solving Bayesian networks exactly \citep{Eklof2013}.
	    However, for larger networks the described numerical evaluation is highly efficient and produces close to identical results \citep{Haussler2020}.

		% BN limitations
		Modelling species extinctions in food webs using Bayesian networks has two main limitations. 
		First, in Bayesian networks consumers do not affect the extinction probabilities of their resources and as such we can here only capture bottom-up effects of disturbances. 
		However, Bayesian networks has been shown to capture the majority of secondary extinctions produced fully-fledged dynamical models \citep{Eklof2013}.
		\cite{Eklof2013} analyzed which functional form of consumers response to loss of resources that most accurately predicted the secondary extinctions realized in the dynamical model, and we here use that functional form in our analyzes.  
		Moreover, bottom-up effects are often the main force of secondary extinctions in empirical systems. 
		For example, bottom-up effects regulate zoo-plankton communities in freshwater lakes, regulating zoo-plankton communities \citep{li2020bottom} and control abundances and biomass of vertebrate species in several terrestrial systems \citep{Dobson2009food, Mduma1999food, Georgiadis2007}. 
		Second, due to the strict bottom-up approach of Bayesian networks the networks must be acyclic.
		Fortunately, \cite{Allesina2009functional} proved that cyclic and acyclic versions of a network have the same properties in terms of secondary extinctions and the effect on network robustness are very limited. 
		Thus, making the networks acyclic in order to use the Bayesian network approach will not strongly affect extinction scenarios.
		Moreover, the cyclic motifs which are excluded from an acyclic network are rare in empirical networks~\citep{Stouffer2007} and therefore unlikely to strongly affect extinction risk.

@@

        In our baseline scenario, all species had an extinction probability of $\pi = 0.1$. 
		To simulate various threats to basal resources, making some species more vulnerable to extinction, we increase basal species' probability of extinction from $0.10$ to $0.5$, in steps of $0.08$. 
		The highest disturbance level, $\pi = 0.5$, corresponds to basal species having a 50\% risk of going extinct. 
		Consumer species retained $\pi=0.1$ in all cases.

@@

		We generated a realistic set of simulated networks based on the niche model~\citep{Williams2000,Stouffer2007} using the function ``nichemodel'' within the Julia~\citep{Bezanson2017julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw,Delmas2017}. 
		To capture a range of plausible network architectures, we simulated networks with sizes (S) ranging from 50 to 100 species (in steps of 10) and connectances (C) ranging from 0.02 to 0.18 (in steps of 0.04). 
		For full details and a comparison to empirical networks, see \emph{Appendix S2}.
		We removed any biologically unlikely networks ( \emph{Appendix S2}) and replaced them with new simulated networks, repeating the process until we obtained 100 suitable networks in each combination of size and connectance.
		To make the simulated networks compatible with the Bayesian simulation framework, all networks were rendered acyclic following~\citet{Allesina2009functional} and species topologically sorted from lowest to highest trophic level (\emph{Appendix S2}). 
		 These processing steps allow the strictly bottom-up calculation of probability of persistence.

@@

    The authors thank György Barabás for excellent discussion of the ideas in this manuscript. AÅ is supported by the Swedish Research Council for Sustainable Development Formas (\#2015-01262) granted to AE. ARC is supported by a Finnish Academy postdoctoral research grant (\#332999).

@@

 %   \begin{figure}[hb!]
 %      \centering
 %       \includegraphics[width=1.0\textwidth]{manuscript/figures/concept_motifs_ab.png}
 %       \caption{Conceptual figure of a toy food web. \textbf{a)} This web contains examples of the four motifs present in an acyclic Bayesian network: apparent competition (blue), three-species chain (purple), omnivory (green), and direct competition (yellow). These motifs capture different ways in which disturbances to basal resources can propagate through the network. For example, a disturbance to the bottom species in a three-species chain will have a direct effect on the middle species but only indirectly affect the top species while a disturbance to the bottom species in an omnivory motif will have a direct effect on the middle species and direct and indirect effects on the top species. Note that the nodes participate in more motifs than the ones highlighted. \textbf{b)} In the same food web, species with similar degree and trophic level can have different motif participation; here species A (light grey node with links in the same motif colored light grey) participates in only two chains while species B (dark grey node with its direct consumption links colored dark grey) participates in chains more than tree times more frequently. Inserted table shows the number of times A and B, respectively, participate in the four different motifs.}
 %   \label{fig:concept}
 %   \end{figure}

>>>>

    Bottom-up effects of disturbances to basal resources can strongly affect the persistence of consumers, but it is difficult to predict which species will be most affected by increasing disturbance.
    Coarse measures of a species' structural role, like in-degree and trophic level, are known to affect a consumer's vulnerability to bottom-up disturbance.
    We are interested in whether a richer set of species roles based on participation in three-species motifs is also related to persistence, and how these motif-based roles relate to simpler measures.
    We show that a consumer species' participation in different motifs is related to its probability of persistence. 
    Similarly, the overall motif profile of a network is related to the average persistence of consumer species.
    Notably, we found that networks containing a higher proportion of the `omnivory' motif tend to have lower mean persistence following disturbance, while more frequent participation in the omnivory motif was associated with higher species-level probabilities of persistence.
    Additionally, we find that relationships between motif participation and persistence can synthesize the information provided by in-degree and trophic level.
    Motif participation therefore captures important trends in species persistence and provides a rich description of species' structural roles in their communities.

@@

    %\subsection*{1 Why are bottom-up effects important for understanding stability?}
     The plant community is the foundation upon which the myriad species in food webs depend for their survival.
     Disturbances to the plant community have been shown to affect important ecosystem properties such as primary \citep{Hector1999} and secondary production \citep{borer2012plant}, soil respiration and carbon cycling \citep{chen2019plant} and consumer diversity \citep{scherber2010bottom, Baiser2016}.
     As such, the structure of the plant community affects stability across all levels of ecosystem organization, which is crucial for sustaining a healthy energy flow in the ecosystem as a whole~\citep{proulx2010diversity, scherber2010bottom, Rosenblatt2016}. 
     For example, the loss of diversity at the basal level typically causes declines in both abundance and richness of all types of consumers: herbivores, predators, parasitoids, etc. \citep{scherber2010bottom}.

    %\subsection*{2 Global and local network properties are known to affect stability}
    How disturbances, such as population decline or extinction, lead to loss of other species via direct and indirect effects (i.e., secondary extinctions) has long been a vibrant area of research \citep{Santos2021,curtsdotter2011robustness, dunne2009cascading, Eklof2006}.
    Several global (whole-network) as well as local (single-species) properties are known to influence how species respond to disturbances. Two well-known, influential global properties are network size, the number of species present in the network, and network connectance, the number of realized feeding interactions.
    Based on simulations of random networks, large networks were initially thought to be less stable~\citep{May1972}.
    The indisputable presence of large, stable, empirical networks has spurred research into how other structural characteristics, including connectance, may allow large numbers of species to persist~\citep{Dunne2002}.
    Highly-connected networks may be more resistant to secondary extinctions because consumers, on average, have access to a larger number of prey species and can switch from a declining or extinct prey to another resource~\citep{Dunne2002, Eklof2006,Baumgartner2015}. However, a well-connected network also includes more pathways through which a disturbance can spread~\citep{Dunne2004, Vieira2015}. Depending on the strength and arrangements of interactions, this means that species may be impacted through several paths, amplifying the original disturbance and causing more secondary extinctions.
    The net effect of connectance on extinction therefore likely depends on finer-scale (species-level) network structure, as well as the type and severity of disturbance.

    Species-level properties known to affect species' responses to disturbances include trophic level (a measure of the path length between a consumer and a basal resource) and in-degree (number of prey items of a predator). 
    These can also be referred to as local properties of the network.
    Species with high trophic levels (long paths to basal resources) and low in-degree (few prey) are generally at higher risks of secondary extinction \citep{binzer2011susceptibility, Eklof2006}, and disturbances to plant populations cause particularly many secondary extinctions \citep{Curtsdotter2011}. %including secondary extinction due to reduced plant populations.
    However, in-degree and trophic level are both relatively coarse descriptions of a species' role in its food web~\citep{Cirtwill2018FoodWebs}. 
    A more detailed perspective is desirable since, as well as the properties of the consumer itself, a consumer's risk of secondary extinction after a disturbance to plants depends on the extinction risk of its prey.
    For example, a generalist consuming many specialist (and therefore high-risk) prey may be at greater risk of extinction than a specialist which consumes a highly generalist resource that is less likely to go extinct (Fig.~\ref{fig:concept}).

@@

        A middle‐ground approach for simulating secondary extinctions is to use Bayesian networks \citep{Eklof2013,Haussler2020}. 
        In this framework, a consumer's probability of extinction depends on the fraction of resources lost ($f = k/n$ for a species with $n$ resources, of which $k$ have gone extinct) and on a baseline probability of extinction ($\pi$) which captures factors affecting the focal species' extinction probability that are not related to the network structure (e.g., disease, stochastic extinction of small populations).
        Modelling Bayesian networks is fast and computationally efficient, allowing analysis of networks with high species richness \citep{Haussler2020}. 
        This increased speed allows us to simulate a range of disturbance severities rather than single species removals as in many previous studies (e.g.,~\citealp[]{Memmott2004,Staniczenko2010,Dunne2004}).
        Moreover, Bayesian network simulations are less parameter-intensive than dynamical models, meaning that they are less sensitive to assumptions made in the modelling process~\citep{Eklof2013}.

@@

		% how we calculate persistences 
        Beginning with basal resources and working systematically up through the food web, we use Equation \ref{betafunc} to calculate each species $i$'s probability of extinction $P(\lnot i|f)$.
        Using this probability, we then perform Bernoulli trials to determine whether the species goes extinct or not. 
        Species are considered extinct if a random number drawn from a uniform distribution $[0-1]$ is less than $P(\lnot i|f)$.
        These simulated extinctions are used when calculating the fraction of resources lost $f$ for each consumer.
        This allows us to update Equation~\ref{betafunc} with information on extinctions of a focal consumer's resources and simulate bottom-up effects. As this method is probabilistic, we repeat these calculations 100 000 times per network, with unique random draws.
	    For each species, we then define probability of persistence as the fraction of simulations in which the species persists. There are methods for solving Bayesian networks exactly \citep{Eklof2013}.
	    However, for larger networks the described numerical evaluation is highly efficient and produces close to identical results \citep{Haussler2020}.

		% BN limitations
		Modelling species extinctions in food webs using Bayesian networks has two main limitations. 
		First, in Bayesian networks consumers do not affect the extinction probabilities of their resources and as such we can here only capture bottom-up effects of disturbances. 
		However, Bayesian networks has been shown to capture the majority of secondary extinctions produced by fully-fledged dynamical models \citep{Eklof2013}.
		\cite{Eklof2013} analyzed which functional form of consumers' response to loss of resources most accurately predicted the secondary extinctions realized in the dynamical model. 
		We here use that functional form in our analyzes.  
		Moreover, bottom-up effects are often the main force of secondary extinctions in empirical systems. 
		For example, bottom-up effects regulate zooplankton communities in freshwater lakes, regulating zooplankton communities \citep{li2020bottom} and control abundances and biomass of vertebrate species in several terrestrial systems \citep{Dobson2009food, Mduma1999food, Georgiadis2007}. 
		Second, due to the strict bottom-up approach of Bayesian networks, the networks must be acyclic.
		Fortunately, \cite{Allesina2009functional} proved that cyclic and acyclic versions of a network have the same properties in terms of secondary extinctions and the effect on network robustness are very limited. 
		Thus, making the networks acyclic in order to use the Bayesian network approach will not strongly affect extinction scenarios.
		Moreover, the cyclic motifs which are excluded from an acyclic network are rare in empirical networks~\citep{Stouffer2007} and therefore unlikely to strongly affect extinction risk.

@@

        In our baseline scenario, all species had an extinction probability of $\pi = 0.1$. 
		To simulate various threats to basal resources, making some species more vulnerable to extinction, we increase all basal species' probability of extinction from $0.10$ to $0.5$, in steps of $0.08$. 
		The highest disturbance level, $\pi = 0.5$, corresponds to basal species having a 50\% risk of going extinct. 
		Consumer species retained $\pi=0.1$ in all cases.

@@

		We generated a realistic set of simulated networks based on the niche model~\citep{Williams2000,Stouffer2007} using the function ``nichemodel'' within the Julia~\citep{Bezanson2017julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw,Delmas2017}. 
		To capture a range of plausible network architectures, we simulated networks with sizes (S) ranging from 50 to 100 species (in steps of 10) and connectances (C) ranging from 0.02 to 0.18 (in steps of 0.04). 
		For full details and a comparison to empirical networks, see \emph{Appendix S2}.
		We removed any biologically unlikely networks (\emph{Appendix S2}) and replaced them with new simulated networks, repeating the process until we obtained 100 suitable networks in each combination of size and connectance.
		To make the simulated networks compatible with the Bayesian simulation framework, all networks were rendered acyclic following~\citet{Allesina2009functional} and species topologically sorted from lowest to highest trophic level (\emph{Appendix S2}). 
		 These processing steps allow the strictly bottom-up calculation of probability of persistence.

@@

    The authors thank Gy\"{o}rgy Barab\'{a}s for excellent discussion of the ideas in this manuscript. A{\AA} is supported by the Swedish Research Council for Sustainable Development Formas (\#2015-01262) granted to AE. ARC is supported by a Finnish Academy postdoctoral research grant (\#332999).

@@
